---
title: "Reproducible Research: Peer Assessment 1"
author: "Andrew E. Davidson"
date: "Sep 15, 2015"
output: 
  html_document:
    keep_md: true
---
```{r setup, echo=FALSE} 
options(scipen=999)
```

## The data
The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

date: The date on which the measurement was taken in YYYY-MM-DD format

interval: Identifier for the 5-minute interval in which measurement was taken

## Loading and preprocessing the data
The github repo we forked from contains the data file activity.zip. 
The following code will automatically unzip the data file into a subdirectory
if needed. In general to save space we do not want to store the unzipped version 
of the data in git.


```{r}
dataDir <- "./data/"
dataFile <- sprintf("%s%s", dataDir, "activity.csv")
if (!file.exists(dataDir)) { dir.create(dataDir)}
if (!file.exists(dataFile)) {
  unzip("activity.zip", exdir=dataDir)
  dateDownLoaded <- date()
  write(dateDownLoaded, file=sprintf("%s%s", dataDir, "dateDownLoaded.txt"))
}
```

Next we need to load the csv file into R. By default read.csv treats the data column as string
factors. We use as.Date() to convert factors to type date
```{r}
data <- read.csv(dataFile, header=TRUE)
data[[2]] <- as.Date(data[[2]])
```

Check to see if any of the fields are missing (For this part of the assignment we can ignore the missing values in the dataset.
)
```{r}
sum(is.na(data$steps))
sum(is.na(data$date))
sum(is.na(data$interval))
```

Check that the data is ordered by date
```{r}
is.unsorted(data$date)
```

Explore the interval data
```{r}
numIntervalsInDay <- (24 * 60) / 5
numUniqueIntervalsInData <- length(unique(data$interval))
rangeOfInterval <- range(data$interval)
```
there are `r numIntervalsInDay` 5 min. intervals in any gvien day. Our data
contains `r numUniqueIntervalsInData` differnt intervals. The range of interval
value is `r rangeOfInterval`. By visually inspecting the interval values I think
the values are actually some sort of time stamp. I.E. 2355 is military time for 
11:55 pm. We should be able to treat the interval data like a factor

## What is mean total number of steps taken per day?

count the number of steps taken each date
```{r}
stepsPerDay <- aggregate(steps ~ date, data=data, FUN=sum)
```


Create a historgram of the total number of steps taken each day
```{r}
library(ggplot2)
qplot(stepsPerDay$steps, geom="histogram") 
```

Calculate the mean and median number of steps per day 
```{r}
averageStepsPerDay <- round(mean(stepsPerDay$steps), digits=2)
medianStepsPerDay <- median(stepsPerDay$steps)
```
The mean number of steps taken each day is `r averageStepsPerDay`.
The median number of steps taken each day is `r medianStepsPerDay`


## What is the average daily activity pattern?

```{r}
averageStepsByInterval <- aggregate(steps ~ interval, data=data, FUN=mean)
p <- ggplot(data=averageStepsByInterval, aes(x=interval, y=steps, group=1))
p + geom_line()
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
m <- max(averageStepsByInterval$steps)
r <- averageStepsByInterval[averageStepsByInterval$steps==m,]
intervalWithMostNumSteps <- r$interval
```
The interval with the most number of steps is `r intervalWithMostNumSteps`

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
numMissingSteps <- sum(is.na(data$steps))
numMissingDates <- sum(is.na(data$date))
numMissingInterval <- sum(is.na(data$interval))
#data[is.na(data)] <- 0
```
The number of rows missing step data is `r numMissingSteps`

The number of rows missing date data is `r numMissingDates`

The number of rows missing interval data is `r numMissingInterval`

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

My stratey is to loop over all rows in the data frame. If the number of steps
is NA, assign the average number of steps for that interval
```{r fixNA}
fixedData <- read.csv(dataFile, header=TRUE)
fixedData[[2]] <- as.Date(fixedData[[2]])
n <- nrow(data)
for (i in 1:n) {
    if (is.na(fixedData[i,1])) {
        interval <- fixedData[i,3]
        x <- averageStepsByInterval[averageStepsByInterval$interval == interval,]$steps
        # we want to round, steps are ints not reals
        r <- round(x)
        fixedData[i,1] <- r
        #cat('interval:', interval, " x:", x, " r:", r, " new:", fixedData[i,1], "\n")
    }
}

sum(is.na(fixedData$steps))
sum(is.na(fixedData$date))
sum(is.na(fixedData$interval))
```

# AEDWIP


## Are there differences in activity patterns between weekdays and weekends?
