---
title: "Reproducible Research: Peer Assessment 1"
author: "Andrew E. Davidson"
date: "Sep 15, 2015"
output: 
  html_document:
    keep_md: true
---
```{r set up} 
options(scipen=999)
```

## Loading and preprocessing the data
The github repo we forked from contains the data file activity.zip. 
The following code will automatically unzip the data file into a subdirectory
if needed. In general to save space we do not want to store the unzipped version 
of the data in git.


```{r}
dataDir <- "./data/"
dataFile <- sprintf("%s%s", dataDir, "activity.csv")
if (!file.exists(dataDir)) { dir.create(dataDir)}
if (!file.exists(dataFile)) {
  unzip("activity.zip", exdir=dataDir)
  dateDownLoaded <- date()
  write(dateDownLoaded, file=sprintf("%s%s", dataDir, "dateDownLoaded.txt"))
}
```

Next we nned to load the csv file into R. By default read.csv treats the data column as string
factors. We use as.Date() to convert factors to type date
```{r}
data <- read.csv(dataFile, header=TRUE)
data[[2]] <- as.Date(data[[2]])
```

Check to see if any of the fields are missing
```{r}
sum(is.na(data$steps))
sum(is.na(data$date))
sum(is.na(data$interval))
```

We need to clean up the data a little. Many samples are missing data. We'll 
convert all NA fields into zeros
```{r}
data[is.na(data)] <- 0
```

Check that the data is ordered by date
```{r}
is.unsorted(data$date)
```
## What is mean total number of steps taken per day?

count the number of steps taken each date
```{r}
stepsPerDay <- aggregate(steps ~ date, data=data, FUN=sum)
```


Create a historgram of the total number of steps taken each day
```{r}
library(ggplot2)
qplot(stepsPerDay$steps, geom="histogram") 
```

Calculate the mean and median number of steps per day 
```{r}
averageStepsPerDay <- round(mean(stepsPerDay$steps), digits=2)
medianStepsPerDay <- median(stepsPerDay$steps)
```
The mean number of steps taken each day is `r averageStepsPerDay`.
The median number of steps taken each day is `r medianStepsPerDay`


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
